Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia

### 2024-06-28 analysis/paper/paper.qmd
- Chi-square SOM - Final JC revisions

---
title: "Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia"
author:
  - Joana Belmiro:
      orcid: https://orcid.org/0000-0003-4951-2907
      correspondence: "yes"
      email: jfbelmiro@ualg.pt
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Xavier Terradas:
      orcid: https://orcid.org/0000-0002-8000-5607
      institute: 
        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
  - Salvador Dominguez-Bella:
      orcid: https://orcid.org/0000-0003-3892-763X
      institute: 
        - UGEA-PHAM, Departamento de Ciencias de la Tierra, Facultad de Ciencias, Universidad de Cádiz. Puerto Real, Cádiz, Spain
  - João Cascalheira:
      orcid: https://orcid.org/0000-0003-0321-8892
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
title-block-published: "Last updated" 
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
number-sections: true
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/springer-socpsych-author-date.csl" # Insert path for the bib-style
abstract: |
  Analyses of raw materials and the distinction between local/regional and long-distance sources have proven invaluable for understanding the extensive movements, interactions, and social networks during the Upper Palaeolithic (UP) in the Iberian Peninsula. However, unlike other parts of Iberia, research on the management and acquisition of raw materials in the south and west of the Peninsula remains relatively underdeveloped. Despite significant knowledge about the technological practices of Palaeolithic hunter-gatherers from southern Portugal, particularly from studies conducted at the Vale Boi site, there's a noticeable lack of focus on raw materials. This paper presents the first comprehensive characterization of chert raw materials from the Gravettian, Proto-Solutrean, and Solutrean sequences at Vale Boi, using both macroscopic and petrographic techniques. Our study reveals that the majority of chert found at Vale Boi originates locally, within a 20 km radius, while a smaller portion comes from non-local sources, potentially indicating long-distance procurement from central Portugal. Additionally, the identification of cherts resembling those from southern Spain suggests a significant west-east mobility pattern throughout the UP period.
keywords: |
  Upper Paleolithic; Iberian Peninsula; Lithic raw materials; Petrography
---

@@

# Call packages
library(readr)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(scales)
library(janitor)
library(ggpubr)
library(knitr)
library(gtsummary)

@@

# Concatenate UNIT and ID variables in one single variable (e.g. A6-100)
context <- context %>%
  dplyr::rename("UNIT" = "Unit") %>% # Renaming first column
  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite Unit with Id s«using cols_to_concat
  select(ID, Level, Spit, Code) %>% # Selecting important variables
  distinct(ID, .keep_all = TRUE)

xyz <- xyz %>%
  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite Unit with Id s«using cols_to_concat
  filter(Suffix == 0) %>% # To avoid several entries related to several points of the same artefact
  select(ID, X, Y, Z) %>% # Selecting important variables
  distinct(ID, .keep_all = TRUE)

@@

T_dataset <- T_dataset %>% 
  filter(!Level %in% c("2", "3")) %>%
  drop_na(Level) %>% 
  mutate(Level = case_when(Level == "7C" ~ "7",
                           Level == 8 ~ "7", # Collapsing samples from Level 7C and Level 8 into 7, since 8 disappears in field_data
                           Level == "4B" ~ "4", # collapsing samples from 4B to 4 due to small number
                           TRUE ~ Level)) %>% 
  mutate(RMTYPE = case_when(RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
                            RMTYPE == "TYPE 15" ~ "TYPE 3",
                            RMTYPE == "TYPE 7" ~ "TYPE 4",
                            RMTYPE == "TYPE 12" ~ "TYPE 5",
                            RMTYPE == "TYPE 11" ~ "TYPE 6",
                            RMTYPE == "TYPE 13" ~ "TYPE 7",
                            RMTYPE == "TYPE 11F" ~ "TYPE 8",
                            RMTYPE == "TYPE 13B" ~ "TYPE 9",
                            RMTYPE == "Oolitic" ~ "TYPE 10",
                            RMTYPE == "GB" ~ "TYPE 11",
                            TRUE ~ RMTYPE))

@@

Knappable raw materials play a crucial role in understanding the mobility and lifeways of past hunter-gatherers, given the ubiquitous presence of lithic artefacts throughout prehistory. More than mere rocks with which stone tools were produced, the management of these resources is intimately connected to a group’s technological, social, and cultural organization, potentially influencing the group’s overall survivability [@oestmoFormalModelingApproach2017; @binfordOrganizationFormationProcesses1979; @bleedOptimalDesignHunting1986; @bousmanHuntergathererAdaptationsEconomic1993; @torrenceTimeBudgetingHuntergatherer1983; @gouldLithicProcurementCentral1985]. Several key topics in the study of hunter-gatherer behaviour and organization have been explored through raw material analysis, such as modalities of procurement and mobility strategies [@ambrose_social_1990; @binfordOrganizationFormationProcesses1979; @binford_righteous_1985; @gouldEmpiricistStrikesBack1985; @gouldLithicProcurementCentral1985; @kuhnUnpackingReductionLithic1991; @mccall_behavioral_2007], occupation types and their duration [@kuhnUpperPaleolithicRaw2004; @surovellBehavioralEcologyLithic2009], the establishment and dimension of social networks [@whallon_social_2006], as well as exchanges between groups or individuals [@gamblePalaeolithicSocietiesEurope1999]. This is often achieved by systematically characterizing geological and archaeological raw materials and establishing correlations between samples from both origins. The ultimate aim is to pinpoint corresponding sources, enabling a precise differentiation between local, regional, and long-distance raw materials, as well as to examine their distribution within a site over different occupations and periods. These approaches and concepts have also been extensively applied to the study of lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula [@aubry_upper_2016; @aubryEconomyLithicRaw2009; @garcia-rojasGreatStepForward2021; @gomezdesolerPanadellaChertMontmaneu2020; @herrero-alonsoManagementLithicRaw2020; @matias_raw_2016; @ortegaMobilitatDesplacamentsDels2003; @pereiraLithicEconomyTerritory2016; @pereira_raw_2021; @pereira_flint_2022; @ramacciottiMovingLandFirst2022; @sanchezdelatorreNewDataChert2023; @soto_procurement_2016], and contributed to understanding how different groups organized their technology, managed time and risk, and explored the available resources during different chronological periods.

Specifically, analyses of raw materials and the distinction between local/regional and long-distance sources have been instrumental in tracing extensive movements, contacts, and social networks throughout the Upper Palaeolithic (UP). Noteworthy examples include studies in northwestern Iberia (Galicia) during the Solutrean period [@hermidaSilexNWPeninsula2016], northern Iberia (Asturias) during the Magdalenian occupations at Las Caldas [@corchonrodriguezMobiliteTerritoiresRelations2016], northeastern Iberia (Catalunya) during the Aurignacian and Gravettian periods at Arbreda cave [@marreirosExploringLithicVariability2016; @ortegaMobilitatDesplacamentsDels2003], northern Portugal during the Gravettian/Solutrean periods in the Côa Valley [@aubry_upper_2016; @aubryRawMaterialProcurement2004], and inland central Iberia (Guadalajara) during the Proto-Solutrean and Solutrean periods at Peña Capón [@sanchezdelatorreNewDataChert2023]. Although such contacts are commonly recognized and identified in later prehistoric periods [@costa_lithic_2022; @noceteRelacionesCentroPeriferia2005; @rodriguezPrehistoricFlintExploitations2011], the recognition of these connections has broadened our comprehension of the often extensive social networks and contacts during the UP in Iberia, illustrating a vast interconnected territory.

However, in contrast to other areas of the Iberia, and despite the potential of raw material studies to shed light on the lifeways and organizational structures of Palaeolithic hunter-gatherers, research on the management and procurement of raw materials during the UP in southern and western Iberia remains in its initial stages. Specifically, in southern Portugal, even though there is considerable knowledge about the technological organization of Palaeolithic hunter-gatherer groups—largely due to research conducted at the Vale Boi archaeological site [see @bicho_ecodynamics_2013; @cascalheiraHunterGathererEcodynamics2013; @marreirosLithicTechnologyGravettian2015]—studies focusing on raw materials are notably infrequent. This situation is attributed to the absence of a detailed and comprehensive characterization of regional lithic resources in the Algarve region (southern Portugal) and the subsequent creation of a reference collection, which is a critical component for conducting raw material studies. Such efforts are particularly crucial for distinguishing between local and non-local raw materials and for understanding variations within the archaeological record [@pop_simulating_2015; @surovellBehavioralEcologyLithic2009]. Recently, we have made significant strides by introducing a reference collection and establishing a framework for raw material studies in the region, through a detailed macroscopic and petrographic database of chert resources in the Algarve region [@belmiroCreatingFramesReference2023]. This initiative paves the way for systematic investigations into raw material procurement strategies during the Upper Palaeolithic, especially concerning the Vale Boi site.

Vale Boi holds a pivotal role in the study of UP adaptations on the Iberian Peninsula, providing invaluable insights into the subsistence and technological strategies of hunter-gatherer populations [@belmiro_gravettian-solutrean_2021; @bicho_ecodynamics_2013; @cascalheira_2017; @cascalheiraTerritorialityOrganizationTechnology2019; @horta_role_2019; @manneIntensiveSubsistencePractices2012; @marreirosExploringLithicVariability2016; @pereira_territory_2016-1]. The rich archaeological record from the site, indicative of prolonged human occupation, is essential for understanding raw material management practices in UP contexts.

Preliminary studies at Vale Boi, employing macroscopic analysis, identified a predominant use of local cherts from Jurassic limestone formations, complemented by a lesser extent of non-local cherts throughout various UP occupations [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004]. According to Pereira et al. [-@pereira_territory_2016-1], the local cherts likely originated within a 16 km radius, reflecting embedded procurement strategies. In contrast, the presence of non-local cherts points to extended sourcing networks, possibly involving travel to other regions of the Algarve, Alentejo, or southern Spain. The presence of chert from the Cretaceous formations of central Portugal has also been suggested [@bicho_ecodynamics_2013].

These results highlight the potential for systematic raw material studies at the site, employing various methods and approaches to analyze the lithic assemblages from multiple UP occupations. Such research aims to comprehend the behaviours of these hunter-gatherer groups over time: how they navigated and utilized their landscape and lithic resources, their movements across a territory bordered by the ocean to the south and west, a large river to the east and mountain ranges to the north, and the nature of the social networks they established. Vale Boi stands as a key archaeological site for this study since it is currently the only site in southern Portugal with a long-term occupation spanning most of the UP and allowing the exploration of raw material use and mobility trends through time [@cascalheira_2017]. Furthermore, previous studies have highlighted the role of Vale Boi in understanding different patterns of territory exploitation and contact with other regions of the Peninsula throughout the UP. During the Solutrean, Vale Boi has been interpreted as an interface between the several sites of Central Portugal and Southern Spain, making it a key site to identify extended social networks that promoted the exchange of information and culture [@cascalheiraValeBoiAlgarve2013; @cascalheira_2017; @cascalheiraSolutrenseEmPortugal2013].

On the contrary, techno-typological studies from the Gravettian occupations of Vale Boi instead show marked differences between regions and suggest a more limited circulation of people and information in southern Iberia [@marreirosLithicTechnologyVariability2013].

To investigate the applicability of these patterns to the management and circulation of lithic raw materials, and to examine the prevailing hypotheses concerning UP mobility and social network exchanges, this paper presents the first comprehensive characterization of chert raw materials derived from the Gravettian, Proto-Solutrean, and Solutrean (c. 32-19 ka cal BP) sequence at Vale Boi.

@@

Vale Boi is located on the western coast of the Algarve region in southern Portugal, within a small valley that extends southward to the Atlantic coast, approximately 2 km away (@fig-map b). It is bordered by limestone outcrops that form rock shelters facing west and southwest [@bichoRockshelterStudiesSouthwestern2007; @bichoEdgeCaseVale2012; @bicho_ecodynamics_2013; @cascalheiraHunterGathererEcodynamics2013; @cascalheira_2017; @manneIntensiveSubsistencePractices2012; @mannePryingNewMeaning2011]. The site extends for more than 10,000 sq meters along the slope of the valley, through which three main areas have been excavated between 2000 and 2019: Slope, Shelter, and Terrace.

The Terrace encompasses occupations from the UP to the Early Neolithic, resulting in the identification of eight main lithostratigraphic units. The UP sequence includes several occupations attributed to the Gravettian (levels 8 to 6) between 32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between 26 and 24 ka cal BP, and Solutrean (level 4D, 4C, 4C, 4 and 3) between 24 and 20 ka cal BP [@belmiro_gravettian-solutrean_2021; @cascalheiraHunterGathererEcodynamics2013; @cascalheira_2017]. The Shelter area shows four main lithostratigraphic units, with Magdalenian, Solutrean and Gravettian occupations. From these, the Solutrean levels show the most intensive occupation, with three, well-preserved, archaeological horizons (layers C to A) and dated from 24 to 22 ka cal BP [@cascalheiraValeBoiAlgarve2013; @cascalheiraTecnologiaLiticaSolutrense2010; @cascalheira_2017].

```{r}
#| label: tbl-c14
#| tbl-cap: Radiocarbon dates for Vale Boi Upper Paleolithic sequence of the Terrace area and Solutrean of the Shelter area, partially adapted from Cascalheira et al. [-@cascalheira_2017] and Belmiro et al. [-@belmiro_gravettian-solutrean_2021], and including new dates. All dates were calibrated or recalibrated using the IntCal20 and Marine20 curves [@reimerIntCal20NorthernHemisphere2020; @heatonMarine20MarineRadiocarbon2020] in OxCal online v.4.4.4 [@ramseyBayesianAnalysisRadiocarbon2009]. For shells a DR value of 123 ± 58 was used, obtained through the CHRONO Marine Reservoir Database at http://calib.org/marine/.

c14_tbl <- read_csv(here::here('analysis/data/raw_data/C14_VB_calBP.csv'))
c14_20 <- read_csv(here::here('analysis/data/raw_data/IntCal20_Marine20_Oxcal.csv'))
dates_cal <- left_join(c14_tbl, c14_20, by = "Lab-ref")
dates_cal <- dates_cal %>% 
  dplyr::rename("Date Cal BP" = "Concatenate") %>% 
  dplyr::select(Area, Level, Culture, `Lab-ref`, Date, SD, `Date Cal BP`, Material)

kable(dates_cal, "simple")

```{r}
#| label: fig-map
#| fig-cap: Iberian Peninsula map with locations and geological formations mentioned throughout the paper. a) Detail of the central Portugal region where the Cenomanian formations with chert nodules are located. b) Detail of southwestern Portugal, with chert and chalcedony bearing formations, outcrops, and location of Vale Boi. c) Detail of eastern Algarve with chert bearing formations. The figure was produced using ArcGIS Pro 3.2.

@@

To characterize the siliceous raw materials used during the UP occupations at Vale Boi and to identify their geological sources, we focused on chalcedony and chert archaeological materials from the Terrace and Shelter areas. These two areas were chosen since both have detailed spatial information, several radiocarbon dates through different levels, and previous studies providing lithic technology, faunal, and geoarchaeological data. Chert was chosen for this study as it is one of the three main used raw materials at the archaeological site of Vale Boi alongside quartz and greywacke. Other raw materials such as chalcedony, dolerite, and schists have also been identified [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @pereira_territory_2016-1]. Using the field database with all individually coordinated lithic artefacts, we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials. @fig-rm-comparison shows that for the selected levels of the Terrace area (@fig-rm-comparison a), chert represents only \~15-20% of the lithic assemblage, while in the Shelter area (@fig-rm-comparison b), chert represents between \~50-70% of the lithic assemblage from the Solutrean layers A, B, and C. Based on previous lithic studies in the terrace and shelter areas, the remaining 80% of other raw materials are mostly composed of quartz and greywacke [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @marreirosPrimeirasComunidadesHomem2009]. Although systematic raw material research has not been conducted on these materials, according to the regional geological context, they have been interpreted as local and readily accessible in the vicinity of the site [@pereira_territory_2016-1]. Frequently, the quartz and greywacke assemblages include large amounts of shattered and/or unknapped chunks or slabs. Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021; @belmiroProtoSolutreanLithicTechnology2020]. Based on this, despite the low amounts of individually plotted chert artefacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits. The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site [@belmiro_gravettian-solutrean_2021].

>>>>

---
title: "Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia"
author:
  - Joana Belmiro:
      orcid: https://orcid.org/0000-0003-4951-2907
      correspondence: "yes"
      email: jfbelmiro@ualg.pt
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Xavier Terradas:
      orcid: https://orcid.org/0000-0002-8000-5607
      institute: 
        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
  - Salvador Dominguez-Bella:
      orcid: https://orcid.org/0000-0003-3892-763X
      institute: 
        - UGEA-PHAM, Departamento de Ciencias de la Tierra, Facultad de Ciencias, Universidad de Cádiz. Puerto Real, Cádiz, Spain
  - João Cascalheira:
      orcid: https://orcid.org/0000-0003-0321-8892
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
title-block-published: "Last updated" 
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
number-sections: true
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/springer-socpsych-author-date.csl" # Insert path for the bib-style
abstract: |
  Analyses of raw materials and the distinction between local/regional and long-distance sources have proven invaluable for understanding the extensive movements, interactions, and social networks during the Upper Palaeolithic (UP) in the Iberian Peninsula. However, unlike other parts of Iberia, research on the management and acquisition of raw materials in the south and west of Iberia remains relatively underdeveloped. Despite significant knowledge about the technological practices of Palaeolithic hunter-gatherers from Southern Portugal, particularly from studies conducted at the site of Vale Boi, there is a noticeable lack of focus on raw materials management. This paper presents the first comprehensive characterization of chert raw materials from the Gravettian, Proto-Solutrean, and Solutrean occupations at Vale Boi, using both macroscopic and petrographic techniques. Our study reveals that the majority of chert found at Vale Boi originates locally, within a 20 km radius. However, a non-neglectable portion of the chert comes from non-local sources, indicating >200 km raw material circulation from Central Portugal and Southern Spain.
keywords: |
  Upper Paleolithic; Iberian Peninsula; Lithic raw materials; Petrography
---

@@

# Call packages
# This includes all necessary packages to run the code
library(readr)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(scales)
library(janitor)
library(ggpubr)
library(knitr)
library(gtsummary)

@@

# Concatenate UNIT and ID variables in one single variable (e.g. A6-100)
context <- context %>%
  dplyr::rename("UNIT" = "Unit") %>% # Renaming first column from Unit to UNIT to use cols_to_concat
  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
  select(ID, Level, Spit, Code) %>% # Selecting important variables: ID refers to artifact ID; Level refers to archaeologycal layer; Spit refers to artificial levels within an archaeological layer; Code refers to type of artifact (e.g., LITHIC or BONE)
  distinct(ID, .keep_all = TRUE)

xyz <- xyz %>%
  unite(col='ID', cols_to_concat, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
  filter(Suffix == 0) %>% # To avoid several entries related to several points of the same artefact
  select(ID, X, Y, Z) %>% # Selecting important variables
  distinct(ID, .keep_all = TRUE)

@@

T_dataset <- T_dataset %>% 
  filter(!Level %in% c("2", "3")) %>% # Filtering out Upper levels
  drop_na(Level) %>% # Dropping artifacts without level
  mutate(Level = case_when(Level == "7C" ~ "7",
                           Level == 8 ~ "7", # Collapsing samples from Level 7C and Level 8 into 7, since 8 disappears in field_data
                           Level == "4B" ~ "4", # collapsing samples from 4B to 4 due to small number
                           TRUE ~ Level)) %>% 
  mutate(RMTYPE = case_when(RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
                            RMTYPE == "TYPE 15" ~ "TYPE 3",
                            RMTYPE == "TYPE 7" ~ "TYPE 4",
                            RMTYPE == "TYPE 12" ~ "TYPE 5",
                            RMTYPE == "TYPE 11" ~ "TYPE 6",
                            RMTYPE == "TYPE 13" ~ "TYPE 7",
                            RMTYPE == "TYPE 11F" ~ "TYPE 8",
                            RMTYPE == "TYPE 13B" ~ "TYPE 9",
                            RMTYPE == "Oolitic" ~ "TYPE 10",
                            RMTYPE == "GB" ~ "TYPE 11",
                            TRUE ~ RMTYPE))

@@

Knappable raw materials play a crucial role in understanding the mobility and lifeways of past hunter-gatherers, given the ubiquitous presence of lithic artefacts throughout Prehistory. More than mere rocks with which stone tools were produced, the management of these resources is intimately connected to a group’s technological, social, and cultural organisation, potentially influencing the group’s overall survivability [@oestmoFormalModelingApproach2017; @binfordOrganizationFormationProcesses1979; @bleedOptimalDesignHunting1986; @bousmanHuntergathererAdaptationsEconomic1993; @torrenceTimeBudgetingHuntergatherer1983; @gouldLithicProcurementCentral1985].

Several key topics in the study of hunter-gatherer behaviour and organization have been explored in different geographic regions and chronological periods through raw material analysis, such as modalities of procurement and mobility strategies [@ambroseSocialEcologicalModels1990; @binfordOrganizationFormationProcesses1979; @binford_righteous_1985; @gouldEmpiricistStrikesBack1985; @gouldLithicProcurementCentral1985; @kuhnUnpackingReductionLithic1991; @mccallBehavioralEcologicalModels2007], occupation types and their duration [@kuhnUpperPaleolithicRaw2004; @surovellBehavioralEcologyLithic2009], the establishment and dimension of social networks [@whallonSocialNetworksInformation2006], as well as exchanges between groups or individuals [@gamblePalaeolithicSocietiesEurope1999]. These analyses are often achieved by systematically characterizing geological and archaeological raw materials and establishing correlations between samples from both origins. The ultimate aim is to identify the corresponding sources, enabling, for example, a precise differentiation between local and non-local raw materials, and examine their distribution within a site over different occupations and periods.

These approaches and concepts have also been extensively and successfully applied to the study of lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula [e.g., @aubryUpperPalaeolithicLithic2016; @aubryEconomyLithicRaw2009; @garcia-rojasGreatStepForward2021; @gomezdesolerPanadellaChertMontmaneu2020; @herrero-alonsoManagementLithicRaw2020; @matiasRawMaterialSourcing2016; @ortegaMobilitatDesplacamentsDels2003; @pereiraLithicEconomyTerritory2016; @pereiraRawMaterialProcurement2021; @pereira_flint_2022; @ramacciottiMovingLandFirst2022; @sanchezdelatorreNewDataChert2023; @sotoProcurementMobilityLate2016; @rodriguezPrehistoricFlintExploitations2011; @costa_lithic_2022; @noceteRelacionesCentroPeriferia2005], and contributed to understanding how different groups explored and managed the available lithic resources.

In this context, raw material analyses and the distinction between local, regional and long-distance sources have been instrumental in tracing extensive movements, contacts, and social networks throughout the Upper Palaeolithic (UP) throughout Iberia. Noteworthy examples include studies in Northwestern Iberia (Galicia) during the Solutrean period [@hermidaSilexNWPeninsula2016], Northern Iberia (Asturias) during the Magdalenian occupations at Las Caldas [@corchonrodriguezMobiliteTerritoiresRelations2016], Northeastern Iberia (Catalunya) during the Aurignacian and Gravettian periods at Arbreda cave [@marreirosExploringLithicVariability2016; @ortegaMobilitatDesplacamentsDels2003], Northern Portugal during the Gravettian/Solutrean periods in the Côa Valley [@aubryUpperPalaeolithicLithic2016; @aubryRawMaterialProcurement2004], and inland Central Iberia (Guadalajara) during the Proto-Solutrean and Solutrean periods at Peña Capón [@sanchezdelatorreNewDataChert2023]. Ultimately, the recognition of these connections has broadened our comprehension of mobility and social organization during the UP in Iberia, illustrating a vast interconnected territory.

However, in contrast to other areas of Iberia, and despite the potential of raw material studies to shed light on the lifeways and organisational structures of Palaeolithic hunter-gatherers, research on the procurement and management of raw materials during the UP in Southern and Western Iberia remains in its initial stages. Specifically, in Southern Portugal, even though there is considerable knowledge about the technological organisation of Palaeolithic hunter-gatherers—largely due to research conducted at the Vale Boi archaeological site [see e.g., @belmiro_gravettian-solutrean_2021; @bicho_ecodynamics_2013; @cascalheiraHunterGathererEcodynamics2013; @marreirosLithicTechnologyGravettian2015; @cascalheiraTecnologiaLiticaSolutrense2010; @horta_role_2019]—raw material studies are notably infrequent and use mainly macroscopic methodologies.

This situation is attributed to the absence of a detailed and comprehensive characterization of regional lithic resources in the Algarve region (Southern Portugal) and the subsequent lack of a complete reference collection produced by complementing characterization methodologies, which is a critical component for conducting raw material studies. Such efforts are particularly crucial for distinguishing between local and non-local raw materials and for understanding variations within the archaeological record [@pop_simulating_2015; @surovellBehavioralEcologyLithic2009].

Recently, we have made significant strides by introducing such a reference collection and establishing a framework for raw material studies in the region, through a detailed macroscopic and petrographic database of chert resources in the Algarve region [@belmiroCreatingFramesReference2023]. This initiative paves the way for systematic investigations into raw material procurement strategies during the Upper Palaeolithic, especially concerning the Vale Boi site, where previous raw material studies have already shown complex patterns of chert procurement and management. These preliminary studies at Vale Boi, employing only macroscopic methods, pointed to a predominant use of local cherts [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004], likely originating within a 16 km radius and reflecting embedded procurement strategies [@pereira_territory_2016-1]. To a lesser extent, non-local cherts were also identified throughout the various UP occupations [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004], pointing to extended sourcing networks, possibly involving sourcing from other regions of Southern Portugal and Southern Spain. The presence of chert from the Cretaceous formations of Central Portugal has also been suggested [@bicho_ecodynamics_2013].

These results highlight the potential for systematic raw material studies at the site to shed light on the lithic resource procurement strategies, mobility patterns, and social networks of the hunter-gatherer groups of Southwestern Iberia throughout the UP.

Vale Boi stands as a key archaeological site for this study since it is currently the only site in Southern Portugal with a long-term occupation spanning most of the UP and allowing the exploration of raw material procurement and mobility trends through time [@cascalheira_2017]. It also holds a pivotal role in the study of UP adaptations on the Iberian Peninsula, providing invaluable insights into the subsistence and technological strategies of hunter-gatherer populations [@belmiro_gravettian-solutrean_2021; @bicho_ecodynamics_2013; @cascalheira_2017; @cascalheiraTerritorialityOrganizationTechnology2019; @horta_role_2019; @manneIntensiveSubsistencePractices2012; @marreirosExploringLithicVariability2016; @pereira_territory_2016-1].

Furthermore, previous studies have highlighted the role of Vale Boi in understanding different patterns of territory exploitation and contact with other regions of Iberia throughout the UP. During the Solutrean, Vale Boi has been interpreted as an important connection point between the core Solutrean areas of Central Portugal and Southern Spain, making it a key site for identifying extended social networks that promoted the exchange of information and culture [@cascalheira_2017; @cascalheiraHunterGathererEcodynamics2013; @cascalheiraSolutrenseEmPortugal2013].

On the contrary, techno-typological studies from the Gravettian occupations of Vale Boi instead show marked differences between regions and suggest a more limited circulation of people and information in Southern Iberia [@marreirosLithicTechnologyVariability2013].

Based on these interpretations and studies, a set of hypotheses and expectations regarding chert procurement and human mobility throughout the UP at Vale Boi can be proposed. On one hand, it would be expected that Solutrean occupations are characterized by the predominant use of local raw materials but with a considerable percentage of cherts obtained from long-distance sources, possibly through exchange with different groups in Iberia. Conversely, it would be expected that Gravettian occupations would be mostly composed of local cherts, with a more limited presence of long-distance cherts.

To investigate the applicability of these patterns to the procurement and circulation of lithic raw materials, and to examine the prevailing hypotheses concerning UP mobility and social network exchanges, this paper focuses on the following questions: 1) what types of chert were the hunter-gatherers of Vale Boi using and where did they come from?; 2) are there different patterns of chert procurement throughout time?; 3) and if so, how can these be related to patterns of land use and the circulation of people/ideas, as suggested through technological studies of the same lithic assemblages? As a result, our study presents the first comprehensive characterisation of chert raw materials, through macroscopic and microscopic analysis, of UP occupations in Southwestern Iberia, derived from the Gravettian, Proto-Solutrean, and Solutrean (c. 32-19 ka cal BP) occupations at Vale Boi.

@@

Vale Boi is located on the western coast of the Algarve region in southern Portugal, within a small valley that extends southward to the Atlantic coast, approximately 2 km away (@fig-map b). It is bordered by limestone outcrops that form rock shelters facing west and southwest [@bichoRockshelterStudiesSouthwestern2007; @bichoEdgeCaseVale2012; @bicho_ecodynamics_2013; @cascalheiraHunterGathererEcodynamics2013; @cascalheira_2017; @manneIntensiveSubsistencePractices2012; @mannePryingNewMeaning2011]. The site extends for more than 10,000 sq meters along the slope of the valley, through which three main areas were excavated between 2000 and 2019: Slope, Terrace, and Shelter. For this study, the chosen areas were the Terrace and Shelter areas (see Section 2 Materials and Methods).

The Terrace encompasses occupations from the UP to the Early Neolithic, resulting in the identification of eight main lithostratigraphic units. The UP sequence includes several occupations attributed to the Gravettian (levels 8 to 6) between c. 32 and 27 ka cal BP, Proto-Solutrean (levels 5 to 4E) between c. 26 and 24 ka cal BP, and Solutrean (level 4D, 4C, 4C, 4 and Lower 3) between c. 24 and 20 ka cal BP [@belmiro_gravettian-solutrean_2021; @cascalheiraHunterGathererEcodynamics2013; @cascalheira_2017].

The Shelter area shows four main lithostratigraphic units, with Magdalenian, Solutrean and Gravettian occupations. From these, the Solutrean levels show the most intensive occupation, with three, well-preserved, archaeological horizons (layers C to A) and dated to c. 24-22 ka cal BP [@cascalheiraValeBoiAlgarve2013; @cascalheiraTecnologiaLiticaSolutrense2010; @cascalheira_2017].

```{r}
#| label: fig-map
#| fig-cap: Iberian Peninsula map with locations and geological formations mentioned throughout the paper. a) Detail of the Central Portugal region where the Cenomanian formations with chert nodules are located. b) Detail of Southwestern Portugal, with chert and chalcedony bearing formations, outcrops, and location of Vale Boi. c) Detail of Eastern Algarve with chert-bearing formations. The figure was produced using ArcGIS Pro 3.2.

@@

To characterize the siliceous raw materials used during the UP occupations at Vale Boi and to identify their geological sources, we focused on chalcedony and chert archaeological materials from the Terrace and Shelter areas. These two areas were chosen since both have detailed spatial information, well-defined chronological sequences, and previous studies providing lithic technology, faunal, and geoarchaeological data.

In this study, the terms chert and chalcedony follow the definitions provided by Luedtke (1992). We use the term chert to refer to sedimentary rocks composed primarily of microcrystalline quartz, which includes other possible varieties such as jasper, agate, or chalcedony. The term chalcedony is used in its petrographic sense and refers to cherts with a fibrous quartz structure.

Chert was chosen for this study as it is one of the three main used raw materials at the archaeological site of Vale Boi alongside quartz and greywacke. Other raw materials such as chalcedony, dolerite, and schists have also been identified [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @pereira_territory_2016-1].

Using the field database with all individually coordinated lithic artefacts (all pieces larger than \~2 cm), we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials. @fig-rm-comparison shows that for the selected levels of the Terrace area (@fig-rm-comparison a), chert represents only \~15-20% of the lithic assemblage, while in the Shelter area (@fig-rm-comparison b), chert represents \~50-70% of the lithic assemblage from the Solutrean layers A, B, and C. Based on previous lithic studies in the Terrace and Shelter areas, the remaining 80% of other raw materials are mostly composed of quartz and greywacke [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @marreirosPrimeirasComunidadesHomem2009].
